### 2023-05-21 bachelorthesis/01-dedication.tex
#39: dedication with picture in black and white
#39: Correction by mayself, #37: Whitspace between formula and sign

\begin{center}
    \vspace*{3cm}
    \begin{tikzpicture}
        \clip(0,0)circle[x radius=3cm, y radius=4cm];
        \node{\includegraphics[width=6cm]{blacky1_black_white.jpg}};
    \end{tikzpicture}
    \vspace{2cm}\\
    This thesis is dedicated to my cat \textbf{Blacky}\\ which is to this date 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ I will miss you everyday when you are gone.\\
    \bigskip
    Manuel Lippert\\
    \today
\end{center}

>>>>

\begin{center}
    \vspace*{3cm}
    \begin{tikzpicture}
        \clip(0,0)circle[x radius=3cm, y radius=4cm];
        \node{\includegraphics[width=6cm]{Dedication/blacky1_black_white.jpg}};
    \end{tikzpicture}
    \vspace{2cm}\\
    (\textborn) 29.05.2005 - (\textdied) 12.05.2023\\
    \bigskip
    This thesis is dedicated to my cat \textbf{Blacky}\\ which get nearly 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ 
    \bigskip
    I will miss you.\\
    \bigskip
    Manuel Lippert\\
    \today
\end{center}

### 2023-05-21 bachelorthesis/02-acknowledgement.tex
#39: Removed name from dedication
#39: Correction by mayself, #37: Whitspace between formula and sign

First of all, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath. I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helpful hands during the the progress of this thesis and the restart script. \bigskip

@@

I would also like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend. In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time sharing the office together. \bigskip

I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling. \bigskip

Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and my girlfriend for their encouragement, endurance and support during the time of this thesis. I know sometimes I am an workaholic and I need to be reminded outside my academic work.

>>>>

First, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath. I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helping hands during the progress of this thesis and the restart script. \bigskip

@@

Also, I would like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend. In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time-sharing the office together. \bigskip

I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get good at english spelling. \bigskip

Outside academia, I would like to express my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support during the time of this thesis. I know sometimes I am a workaholic and I need to be reminded about a life outside my academic work.

### 2023-05-21 bachelorthesis/03-abstract.tex
#39: Fixed Linebreak
#39: Correction by mayself, #37: Whitspace between formula and sign

Ion temperature gradient driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures. Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies.

To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations.

For convenience a \texttt{python} script (\texttt{slurm\_monitor.py}) was written which monitors the simulation on the \texttt{btrzx1}-cluster and starts/restarts until the completion criteria is fulfilled.

Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit. This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG driven turbulence.

@@

Ionen-Temperaturgradienten getriebener Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf. Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt.

Um den Aufwand der Berechnungen der Simulationen zu reduzieren wurden mehrere kleinere Inputparameter für GKW (Gyro Kinetic Workshop) getestet um die optimal kleinste Auflösung zu finden ausgeführten Simulationen zu finden.

Zur Einfachheit wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist.

Weiterhin wurde doch mehrere radiale konvergierende Boxgrößen-Scans gezeigt, dass eine Mesoskalengröße von $\sim 57.20-76.27\,\rhoth$ inhärent zur ITG getriebenen Turbulenz mit Cyclone Base Parameter in lokalen Limit ist. Dieses Ergebnis impliziert auch, dass eine typische Skala für den lawinenenartig Transport inhärent für die ITG getriebene Turbulenz ist.

>>>>

Ion temperature gradient driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures. Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies. \bigskip

To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations. \bigskip

For convenience a \texttt{python} script (\texttt{slurm\_monitor.py}) was written which monitors the simulation on the \texttt{btrzx1}-cluster and starts/restarts until the completion criteria is fulfilled. \bigskip

Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57-76\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit. This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG driven turbulence.

@@

Ionen-Temperaturgradienten getriebene Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf. Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt. \bigskip

Um den Aufwand der Berechnungen der Simulationen zu reduzieren wurden mehrere kleinere Inputparameter für GKW (Gyro Kinetic Workshop) getestet um die optimal kleinste Auflösung für die ausgeführten Simulationen zu finden. \bigskip

Zur Einfachheit wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist. \bigskip

Weiterhin wurde doch mehrere radiale konvergierende Boxgrößen-Scans gezeigt, dass eine Mesoskalengröße von $\sim 57-76\,\rhoth$ inhärent zur ITG getriebenen Turbulenz mit Cyclone Base Parameter in lokalen Limit ist. Dieses Ergebnis impliziert auch, dass eine typische Skala für den lawinenenartig Transport inhärent für die ITG getriebene Turbulenz ist.

### 2023-05-21 bachelorthesis/04-declaration.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. This thesis is based on
\bigskip
\begin{itemize}
    \item [] \bibentry{Lippert2023}
\end{itemize}
\bigskip
and is a further iteration of this publication (brief communication published in the journal \enquote{Physics of Plasma}). It provides additional plots and paragraphs that were not necessary to be included in the final publication. The format was changed as well because the publication was written in REV\TeX~4.1 and this bachelor thesis in \texttt{scrreprt}. The brief communication can be found in Appendix \ref{append:paper}.

>>>>

The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. This thesis is based on
\bigskip
\begin{itemize}
    \item [] \bibentry{Lippert2023}
\end{itemize}
\bigskip
and is a further iteration of this publication (brief communication). It provides additional plots and paragraphs that were not necessary to be included in the final publication. The brief communication can be found in Appendix \ref{append:paper}.

### 2023-05-21 bachelorthesis/21-motion.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under tha associated force. Charged particles can be separated in species, e.g. electrons and ions, which will be later on not displayed in the governing equation. Throughout this thesis the charge $q$ or the mass $m$ indicate the quantities of a specific spiecies.

@@

Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as:
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by:
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
which center is often as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.

Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical species thermal velocity is
\begin{gather}
    \vth = \sqrt{\frac{2T}{m}}~,
\end{gather}
where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as \cite{Wesson2011}
\begin{gather}
    \rhoth = \frac{m\vth}{|q|B}~.
\end{gather}

@@

When an electric field with a component parallel to the magnetic field $E_\parallel$ acts on plasma the charged particles gets accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = qE_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.

Since magnetic fields are not always homogenous an inhomogeneous magnetic field with its gradient $\nabla B$ contains a component parallel to the magnetic field which is given by:
\begin{gather}
    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B~.
\end{gather}
This causes the force
\begin{gather}
    F_{\parallel,\nabla_\parallel B} = - \frac{mv^2_{\perp}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$ which can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The caused force has it application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

\begin{enumerate}
    \item \textbf{$\exb$ Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component is present together with the magnetic field $\vect{B}$ (both fields are homogenous) the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$-drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item \textbf{$\nabla B$ Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$-drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The $\nabla B$-drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$-drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic filed the gyrocenter experience a curvature drift given by
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small in comparison die the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

>>>>

In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under the associated force. Charged particles can be separated in species, e.g. electrons and ions, which will be later on not displayed in the governing equation. Throughout this thesis the charge $q$ the mass $m$ or the temperature $T$ indicate the quantities of a specific species.

@@

Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
which center is often referred as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.\\\bigskip

Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical thermal velocity is
\begin{gather}
    \vth = \sqrt{\frac{2T}{m}}~,
\end{gather}
where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as
\begin{gather}
    \rhoth = \frac{m\vth}{|q|B}~. \cite{Wesson2011}
\end{gather}

@@

When an electric field with a component parallel to the magnetic field $E_\parallel$ acts on plasma the charged particles gets accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = qE_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.\\\bigskip

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
and causes the force
\begin{gather}
    F_{\parallel,\nabla_\parallel B} = - \frac{mv^2_{\perp}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The caused force has it application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

\begin{enumerate}
    \item \textbf{$\exb$ Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item \textbf{$\nabla B$ Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The $\nabla B$ drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic filed the gyrocenter experience a curvature drift given by
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small in comparison die the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

### 2023-05-21 bachelorthesis/22-confinement.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

In a tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion containing gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R_0$ the particle and energy losses are caused by the gyro center drift. To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux sufaces} and with magnetic field lines which appearing on that surfaces. For stability reasons the magnetic field has a toroidal and a poloidal component which also holds equivalence of the plasma pressure. \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current. Together the components results in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}]. To characterize the the quality of confinement the so-called \textit{plasma beta} is used and is given by
\begin{gather}
    \beta = \frac{nT}{\mu_0 B^2/2}~.
\end{gather} 
with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$. Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates when the plasma beta is smaller the confinement is better. In a tokamak reactor the plasma beta has a typical order of few percent. \cite{Wesson2011}
\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or zylinder coordinates ($Z$, $R$, $\phi$). \cite{Barton2015}
}{fig:confinement}{0.7}

>>>>

In tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion which contains the gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift. To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux sufaces} and with magnetic field lines which appearing on that surfaces. For stability reasons the magnetic field has a toroidal and a poloidal component which also holds equivalence of the plasma pressure. \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current. Together the components results in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}]. To characterize the the quality of confinement the so-called \textit{plasma beta} is used and is given by
\begin{gather}
    \beta = \frac{nT}{\mu_0 B^2/2}~.
\end{gather} 
with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$. Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates when the plasma beta is smaller the confinement is better. In a tokamak reactor the plasma beta has a typical order of few percent. \cite{Wesson2011}
\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or zylinder coordinates ($Z$, $R$, $\phi$). \cite{Barton2015}
}{fig:confinement}{0.7}

### 2023-05-21 bachelorthesis/23-iontempgrad.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

The described electrostatic forces acting on the tokamak plasma causes a variety of collective phenomena. Microinstabilities is one of this phenomena which are instabilities on the spatial scale of the Larmor radius and frequencies much smaller than the cyclotron frequency. These instabilities are driven bei density and temperature gradients that are present in the tokamak devices. Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors. The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability \cite{Coppi1967, Cowley1991, Rudakov1961}

\includegraphicsHere{Theory/ITG-Turbulence.pdf}{
	Sketch of ion temperature gradient driven instability on the outboard low field sid of the tokamak \cite{Casson_PHD}
}{fig:ITG}{0.5}

The ITG driven is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard side of the tokamak. The mechanism will be briefly explained with the help of Fig. \ref{fig:ITG} \cite{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature and $\nabla B$ dirft $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential causes an electric field which results in a $\exb$ drift $\vect{v}_E$. The $\exb$ drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore such turbulent structures are called \textit{eddies} and the have a typical radial scale of serval Larmor radius. \cite{Newins2006}. 
Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

>>>>

The described electrostatic forces acting on the tokamak plasma causes a variety of collective phenomena. Microinstabilities is one of these phenomena which are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$. These instabilities are driven by density and temperature gradients that are present in the tokamak devices. Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors. The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability. \cite{Coppi1967, Cowley1991, Rudakov1961}

\includegraphicsHere{Theory/ITG-Turbulence.pdf}{
	Sketch of ion temperature gradient driven instability on the outboard low field side of the tokamak \cite{Casson_PHD}
}{fig:ITG}{0.5}

The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism will be briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ dirft with velocity $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of serval Larmor radius. \cite{Newins2006}
Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

### 2023-05-21 bachelorthesis/24-zonalflows.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

Zonal flows are referred in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}. \cite{Diamond2005} Because of the variation only across flux-surfaces only the zonal flow is tangential to the flux-surfaces and do not contribute to the turbulent transport across flux-surfaces caused by the $\exb$ motion. 
Zonal flows are driven nonlinear through turbulent Reynolds stresses \cite{Diamond1991}. For more about the mechanism behind zonal flow generation the reader is referred to Ref. \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}. \bigskip

@@

When the zonal flows are strong enough to suppress turbulence completely due to the shearing process, pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. The $\exb$ staircase pattern has the following properties:
\begin{enumerate}
	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10 \rhoth which is larger than the Larmor radius but smaller than the machine size $R_0$.
	\item[(2)] The structure are \textit{quasi stationary} in space and varies on time scales much longer than the typical turbulent time scales.
	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$-shearing rate of the order of $10^{-1} \vth/R_0$.
	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
\end{enumerate} \bigskip

The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
\end{equation}
The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}. \\ \bigskip

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing \cite{Idomura2009,McMillan2009}. The fully developed avalanche resluts in almost completely quenched turbulence. Its development results in a discontinuous step int he dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}

>>>>

Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}. \cite{Diamond2005} Because of the variation only across flux-surfaces the zonal flow is tangential to those surfaces and do not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion. 
Zonal flows are driven nonlinear through turbulent Reynolds stresses. \cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref. \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}. \bigskip

@@

When the zonal flows are strong enough to suppress turbulence completely due to the shearing process, pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. The $\exb$ staircase pattern has the following properties:
\begin{enumerate}
	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10\,$\rhoth$ which is larger than the Larmor radius but smaller than the machine size $R$.
	\item[(2)] The structure are \textit{quasi stationary} in space and varies on time scales much longer than the typical turbulent time scales.
	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$ shearing rate of the order of $10^{-1} \vth/R$.
	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
\end{enumerate} \bigskip

The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}~,
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)~.
\end{equation}
The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}. \\ \bigskip

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing. \cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}
